% Mit twoside als weitere Option werden die Seitenränder angepasst, sodass beidseitig gedruckt werden kann
\documentclass[pdftex,12pt,a4paper]{report}

% kann optional mit "german" als Sprache gewählt werden
\usepackage[]{dbstmpl}



\global\arbeit{Bachelor Thesis}
%\global\arbeit{Masterarbeit}
%\global\arbeit{Seminararbeit}




% Hier die eigenen Daten eintragen
\global\titel{Action-Head Learning on Frozen Vision-Language Models for an Autonomous Minecraft Agent}
\global\bearbeiter{Diego Cuenca}
\global\betreuer{Prof. Volker Tresp}
\global\aufgabensteller{AUFGABENSTELLERNAME}
\global\abgabetermin{30.06.2026}
\global\ort{Munich}
\global\fach{Computer Science}

\begin{document}

% Deckblatt
\deckblatt

% Erklaerung fuer das Pruefungsamt
% Bei Seminararbeiten nicht noetig
\erklaerung

% Zusammenfassung
\begin{abstract}
Vision-language models carry broad visual and semantic priors, which makes them
attractive backbones for embodied agents --- but turning one into an agent
usually means post-training the whole model on action data, tying it to a single
embodiment at substantial cost. This thesis asks how far the cheap alternative
goes: freeze the backbone, train only a small ($\sim$5M-parameter) action head by
behavioural cloning, and keep the whole study inexpensive enough to sweep on a
single rented GPU. Using Minecraft as a deliberately hard test bed, it runs a
$2\times2$ ablation over backbone (the joint LLaVA-1.5 versus the contrastive
CLIP) and language condition (task prompt versus none), evaluating every cell
both on held-out frames and in live MineRL rollouts.

Three findings emerge, each replicated across three training seeds. First, under
the minimal concatenating head that the controlled comparison holds fixed, the
contrastive backbone is the more effective controller --- CLIP collects dirt where
the matched LLaVA cell sits at the floor --- but this is a property of the head
read-out, not of the backbone: two alternative heads recover CLIP-comparable
collection from the same frozen LLaVA features, making the result a head-by-backbone
interaction rather than an intrinsic ranking. Second, language conditioning is
likewise a backbone effect --- only CLIP's separate text encoder yields a prompt
signal the head can act on (a $42$-point attack-rate shift, large in every seed),
while LLaVA stays flat across prompts even though the prompt is almost perfectly
decodable from its features. Third, and most robustly, these differences are
invisible offline: the four cells are a near-tie on frame metrics yet diverge from
a dug stack of dirt to nothing per episode in rollout --- a direct demonstration
that closed-loop evaluation measures what frame-level metrics miss. The study
characterises the cheap-and-frozen corner of the design space and does not claim
either backbone's ceiling under fine-tuning.
\end{abstract}

% Inhaltsverzeichnis (abhaengig von der Laenge des Dokuments)
\tableofcontents

% Hier beginnt der Hauptteil

\input{chapter1}
\input{chapter2}
\input{chapter3}
\input{chapter4}
\input{chapter5}
\input{chapter6}







\newread\tmp

\newcommand{\quickcharcount}[1]{%
  \immediate\write18{texcount -1 -sum -merge -char #1.tex > #1-chars.sum}%
  \openin\tmp=#1-chars.sum%
  \read\tmp to \thechar%
  \closein\tmp%
}

\newcommand{\quickwordcount}[1]{%
  \immediate\write18{texcount -1 -sum -merge #1.tex > #1-words.sum}%
  \openin\tmp=#1-words.sum%
  \read\tmp to \theword%
  \closein\tmp%
}


\quickwordcount{main}
\quickcharcount{main}

There are \thechar characters and approximately \theword spaces.
That makes approximately \the\numexpr\theword+\thechar\relax\ characters total.


%% Ende Hauptteil

% Abbildungsverzeichnis (kann auch nach dem Inhaltsverzeichnis kommen)
%\listoffigures

% Tabellenverzeichnis (kann auch nach dem Inhaltsverzeichnis kommen)
%\listoftables

% Literaturverzeichnis
\bibliographystyle{plain}
\bibliography{dbstmpl}


% verwendet dbstmpl.bib

\end{document}
